State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-11-29 manuscript/manuscript_2/paper.tex
update manuscript for n4000 results

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. It is shown how to extend previous state-space analyses to include the pulsar terms. Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%. For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated. However, including the pulsar terms decreases the minimal detectable strain by \textcolor{red}{Y percent} .
\end{abstract}

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ , \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling, and the combined workflow in the context of PTA searches for GWs. \newline

@@

			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)_{\rm ATNF}} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\

@@

		\caption{Summary of injected static parameters used for generating synthetic data for validation of the analysis scheme when including the pulsar-term in Section \ref{sec:rep_example}. The prior used for Bayesian inference (Section \ref{sec:pe_and_ms}) on each parameter is also displayed (rightmost column).  The top, middle and bottom sections of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$, $\boldsymbol{\theta}_{\rm psr}$ and $\chi^{(n)}$ respectively. The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synthetic_data}. The subscript ``SC'' on the pulsar noise amplitude $\sigma^{(n)}$ indicates that the injected value has been calculated using Equation and \eqref{eq:sigmap_f} the empirical timing noise model for MSPs from \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. We do not try to infer $\gamma^{(n)}$ as $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$ and so no prior is set. The choice of priors on $\boldsymbol{\theta}_{\rm psr}$ is discussed in Appendix \ref{sec:set_priors}.

@@

In order to generate synthetic frequency time series $ \boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$. The measured $f_{\rm m}^{(n)}(t)$ are then produced via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:sigma_m_eqn}). We now describe how $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ are chosen. This is equivalent to specifying the configuration of a synthetic PTA. In Section  \ref{sec:pe_and_ms} and Section \ref{sec:parameter_space} we describe how the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$, are chosen. This is equivalent to specifying the synthetic SMBHB source.  \newline

@@

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for nine arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. Each coloured curve is the result for different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. There is an evident offset in the peaks of the posteriors of $\iota$ and $h_0$ relative to the injection value, across all noise realisations \textcolor{red}{TK: to discuss}. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 45 out of the 63 possible combinations of the seven parameters and 63 realizations (i.e. 70 $\%$). There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations. The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 43\% for $\Phi_0$. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (Figure \ref{fig:corner_high_snr_appendix}). At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value. \newline

In addition to estimates of $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach. Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the nine noise realisations, analogously to Figure \ref{fig:corner_plot_1}. We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big. With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is no evidence for correlations between parameter pairs. The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e. not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 325 out of the 423 possible combinations of the 47 parameters and nine realizations (i.e. $\sim$ 77 $\%$). There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}). This corresponds to the phase-correction term for PSR J0340+4130. This pulsar lies in the closest direction to our synthetic gw source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$. As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the pulsar term is minimal. The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}. \newline

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters. The known, injected value lies within the 90\% credible interval for 45 out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors; the maximum coefficient of variation is $43 \%$ for $\Phi_0$, the minimum is $0.02 \%$ for $\Omega$.}
	\label{fig:corner_plot_1}
\end{figure*}

%images/corner_lowSNR_chiF
	\begin{figure}
	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$. Consistent unimodal posteriors are obtained across the nine noise realisations for four out of the five displayed parameters. No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source. The known, injected value lies within the 90\% credible interval for 37 out of the 45 combinations of five parameters and nine noise realizations (i.e. $82 \%$). }
	\label{fig:corner_plot_3}
\end{figure}

\subsection{Extension to broader parameter space} \label{sec:parameter_space}
%images/pp_plot_new1
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
	\label{fig:parameter_space}
\end{figure}
Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. In this section we test the method for various systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges. The aim of this section is to confirm that the analysis scheme can be applied across an astrophysically relevant parameter space and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Section \ref{sec:pe_and_ms} is not particularly advantageous by accident. \newline

@@

It is prohibitive to produce corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$. Therefore, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself. In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope. The results are displayed in Figure \ref{fig:parameter_space}.  
The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters generally exhibit the discussed linear behaviour. Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2-$\sigma$ shaded region.

@@

\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm P \& E}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm E}$. Model $\mathcal{M}_{\rm P \& E}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm E}$ is parameterised by $\boldsymbol{\theta}_{\rm E} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm P \& E}$, as described in Appendix \ref{sec:set_priors}. \newline

In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm E}$ returned by the respective models. In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable strain for the two models.

@@

In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model. For $\mathcal{M}_{\rm E}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm E})$. The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm E} | \boldsymbol{Y})$. For $\mathcal{M}_{\rm P \& E}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$. The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$. The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models. \newline

@@

Extensions to the state-space methods for PTA data analysis are discussed at length in K24. We note three relevant extensions here:

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. It is shown how to extend previous state-space analyses to include the pulsar term. Astrophysically representative software injections in Gaussian measurement noise evidence that for a loud GW source with characteristic wave strain $h_0 = 10^{-12}$ the inclusion of the pulsar terms decreases the root-mean-square relative error in the estimated one-dimensional posteriors by up to 99\%. For a quiet GW source with $h_0 = 5 \times 10^{-15}$ there is no significant improvement in the parameter accuracy since the system is heavily noise dominated. However, including the pulsar terms decreases the minimal detectable strain by \textcolor{red}{Y percent} .
\end{abstract}

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}, $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$, and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. In the context of PTA searches for GWs, we refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, Appendix \ref{sec:nested_sampling} for a review of nested sampling, Appendix \ref{sec:workflow} for an overview of the combined Kalman filter - nested sampling workflow. \newline

@@

			\vspace{1mm} &  $\chi^{(n)}$       &$\Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}_{\rm ATNF} \right)  d_{\rm ATNF}^{(n)} $  & rad & Uniform($0, 2 \pi $) \\

@@

		\caption{Summary of the injected static parameters used for generating synthetic data for validation of the analysis scheme when including the pulsar-term in Section \ref{sec:pe_and_ms}. The prior used for Bayesian inference (Section \ref{sec:pe_and_ms}) on each parameter is also displayed (rightmost column).  The top, middle and bottom sections of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$, $\boldsymbol{\theta}_{\rm psr}$ and $\chi^{(n)}$ respectively. The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synthetic_data}. The subscript ``SC'' on the pulsar noise amplitude $\sigma^{(n)}$ indicates that the injected value has been calculated using Equation \eqref{eq:sigmap_f} the empirical timing noise model for MSPs from \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. We do not try to infer $\gamma^{(n)}$ as $\gamma^{(n)} T_{\rm obs} \sim 10^{-5}$ and so no prior is set. The choice of priors on $\boldsymbol{\theta}_{\rm psr}$ is discussed in Appendix \ref{sec:set_priors}.

@@

In order to generate synthetic frequency time series $ \boldsymbol{Y}(t) = \{f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) \}$, Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$. The measured $f_{\rm m}^{(n)}(t)$ are then produced via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:vareps}). We now describe how $\boldsymbol{\theta}_{\rm psr}$, ${\boldsymbol{q}}^{(n)}$ and $\sigma_{\rm m}$ are chosen. This is equivalent to specifying the configuration of a synthetic PTA. In Section  \ref{sec:pe_and_ms} and Section \ref{sec:parameter_space} we describe how the remaining elements of $\boldsymbol{\theta}'$, namely $\boldsymbol{\theta}_{\rm gw}$, are chosen. This is equivalent to specifying the synthetic SMBHB source.  \newline

@@

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for ten arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. Each coloured curve is the result for different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations (i.e. 86$\%$). The injected value is contained within the 90\% credible interval in 100\% of cases for $\Omega, \Phi_0, \psi$, and  $\iota$. The injected value is contained within the 5\% and 95\% quantiles for all of the one-dimensional marginalized posteriors.

There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations. The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 52\% for $\Phi_0$. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR}, Figure \ref{fig:corner_high_snr_appendix}. At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value. \newline

In addition to $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach. Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations, analogously to Figure \ref{fig:corner_plot_1}. We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big. With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is no evidence for correlations between parameter pairs. The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e. not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations (i.e. $\sim$ 81 $\%$). There are some elements for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}). This corresponds to the phase-correction term for PSR J0340+4130. This pulsar lies in the closest direction to our synthetic GW source, such that $| \boldsymbol{n} \cdot \boldsymbol{q}^{(2)}| = 0.96$. As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the GW is minimal ($g^{(2)}(t) \to 1$) The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately, see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}. \newline

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR_n4000}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for ten realisations of the noise processes, with curves coloured uniquely. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters. The known, injected value lies within the 90\% credible interval for 60 out of the 70 combinations of seven parameters and ten noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors; the maximum coefficient of variation is $52 \%$ for $\Phi_0$, the minimum is $0.02 \%$ for $\Omega$.}
	\label{fig:corner_plot_1}
\end{figure*}

%images/corner_lowSNR_chiF
	\begin{figure}
	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi_n4000}
	\caption{Same as Figure \ref{fig:corner_plot_1}	but for five static parameters of $\chi^{(n)}$. Consistent unimodal posteriors are obtained across the nine noise realisations for four out of the five displayed parameters. No consistent posteriors are obtained for $\chi^{(2)}$ (third row, third column), because this pulsar has a similar sky position to that of the synthetic GW source. The known, injected value lies within the 90\% credible interval for 38 out of the 50 combinations of five parameters and nine noise realizations (i.e. $76 \%$). }
	\label{fig:corner_plot_3}
\end{figure}

\subsection{Extension to broader parameter space} \label{sec:parameter_space}
%images/pp_plot_new1
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
	\label{fig:parameter_space}
\end{figure}
Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. In this section we test the method for a range of systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges. The aim of this section is to confirm that the analysis scheme can be applied across an astrophysically relevant parameter space and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Section \ref{sec:pe_and_ms} is not particularly advantageous by accident. \newline

@@

It is prohibitive to display corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$. Therefore, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself. In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope. The results are displayed in Figure \ref{fig:parameter_space}.  
The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters generally exhibit the discussed linear behaviour. Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2-$\sigma$ shaded region.

@@

\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline

In Section \ref{sec:psr_v_earth_pe} we compare the accuracy of the estimates of $\boldsymbol{\theta'}$, $\boldsymbol{\theta}_{\rm earth}$ returned by the respective models. In Section \ref{sec:psr_v_earth_bayes} we compare the minimal detectable GW strain for the two models.

@@

In this section apply we apply the Kalman filter in conjunction with nested sampling in order to infer the joint posterior distribution for the static parameters of each model. For $\mathcal{M}_{\rm earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_earth} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}_{\rm earth})$. The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta}_{\rm earth} | \boldsymbol{Y})$. For $\mathcal{M}_{\rm psr \& earth}$ we apply the Kalman filter using Equation \eqref{eq:g_func_trig_chi} to return $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}')$. The likelihood returned by the Kalman filter is then used by nested sampling in order to estimate $p(\boldsymbol{\theta'} | \boldsymbol{Y})$. The settings of the nested sampler (for example the number of live points, see Appendix \ref{sec:nested_sampling}) are identical for both models. \newline

@@

Extensions to the state-space methods for PTA data analysis are discussed at length in K24. We note three relevant extensions here: \newline

(0) there is no real improvement in PE for this particular representative source. Exploration of a broader parameter space.... \textcolor{red}{TK: TBD}
